Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-13 talk/pepm2011/paper.tex
(antocuni, cfbolz) a minimal fix for one of antos remarks that I just killed
some changes to Sect 1

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT works by observing the running
program and recording its hot parts into linear execution traces, which are then turned into
machine code. One reason for the popularity of tracing JITs is their relative
simplicity. They can often be added to an interpreter and a lot of the
infrastructure of an interpreter can be reused. They give some important
optimizations like inlining and constant-folding for free. A tracing JIT always
produces linear pieces of code, which simplifies many optimizations that are usually
hard in a compiler, such as register allocation.

The usage of a tracing JIT can remove the overhead of bytecode dispatch and that
of the interpreter data structures. In this paper we want to present an approach
how an optimization can be added to a tracing JIT that also removes some of the
overhead more closely associated to dynamic languages, such as boxing overhead
and type dispatching. Our experimental platform is the PyPy project, which is an
environment for implementing dynamic programming languages. PyPy and tracing
JITs are described in more detail in Section~\ref{sec:Background}.
Section~\ref{sec:lifetimes} analyzes the problem to be solved more closely.

The most important technique we use to achieve to optimize traces is a
straightforward application of partial evaluation. The partial evaluation
performs a form of escape analysis \cite{bruno_blanchet_escape_2003} on the traces and make some
objects that are allocated in the trace \emph{static}\footnote{These objects are
called \emph{virtual} in Psyco \cite{rigo_representation-based_2004}.} which
means that they do not occur any more in the optimized trace. This technique is
informally described in Section~\ref{sec:statics}, a more formal description is
given in Section~\ref{sec:formal}.

@@

The core idea of tracing JITs is to focus the optimization effort of the JIT
compiler on the hot paths of the core loops of the program and to just use an
interpreter for the less commonly executed parts. VMs that use a tracing JIT are
thus mixed-mode execution environments, they contain both an interpreter and a
JIT compiler. By default the interpreter is used to execute the program, doing
some light-weight profiling at the same time. This profiling is used to identify
the hot loops of the program. If a hot loop is found in that way, the
interpreter enters a special \emph{tracing mode}. In this tracing mode, the
interpreter records all operations that it is executing while running one
iteration of the hot loop. This history of executed operations of one loop is
called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
it always ends with a jump to its own beginning. The trace also contains all
operations that are performed in functions that were called in the loop, thus a
tracing JIT automatically performs inlining.

>>>>

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT works by observing the running
program and recording its hot spots into linear execution traces, which are then turned into
machine code. One reason for the popularity of tracing JITs is their relative
simplicity. They can often be added to an interpreter and a lot of the
infrastructure of an interpreter can be reused. They give some important
optimizations like inlining and constant-folding for free. A tracing JIT always
produces linear pieces of code, which simplifies many optimizations that are usually
hard in a compiler, such as register allocation.

The usage of a tracing JIT can remove the overhead of bytecode dispatch and that
of the interpreter data structures. In this paper we want to present a new
optimization that can be added to a tracing JIT that further removes some of the
overhead more closely associated to dynamic languages, such as boxing overhead
and type dispatching. Our experimental platform is the PyPy project, which is an
environment for implementing dynamic programming languages. PyPy and tracing
JITs are described in more detail in Section~\ref{sec:Background}.
Section~\ref{sec:lifetimes} analyzes the problem to be solved more closely.

The core of our trace optimization technique can be
viewed as partial evaluation: the partial evaluation
performs a form of escape analysis \cite{bruno_blanchet_escape_2003} on the traces and make some
objects that are allocated in the trace \emph{static}\footnote{These objects are
called \emph{virtual} in Psyco \cite{rigo_representation-based_2004}.} which
means that they do not occur any more in the optimized trace. This technique is
informally described in Section~\ref{sec:statics}, a more formal description is
given in Section~\ref{sec:formal}.

@@

The core idea of tracing JITs is to focus the optimization effort of the JIT
compiler on the hot paths of the core loops of the program and to just use an
interpreter for the less commonly executed parts. VMs that use a tracing JIT are
thus mixed-mode execution environments, they contain both an interpreter and a
JIT compiler. By default the interpreter is used to execute the program, doing
some light-weight profiling at the same time. This profiling is used to identify
the hot loops of the program. If a hot loop is found in that way, the
interpreter enters a special \emph{tracing mode}. In this tracing mode, the
interpreter tries to record all operations that it is executing while running one
iteration of the hot loop. This history of executed operations of one loop is
called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
it always ends with a jump to its own beginning. The trace also contains all
operations that are performed in functions that were called in the loop, thus a
tracing JIT automatically performs inlining.
